Geometry of Ranked Nearest Neighbour Interchange Space of Phylogenetic Trees

### 2019-12-11 rnni_convexity.tex
Discussion, sitations, acknowledgements

\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage[UKenglish]{babel}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amssymb}
\usepackage{float}
\usepackage{graphicx}
\usepackage{algorithm}
\usepackage[noend]{algorithmic}
\usepackage{todonotes}
\usepackage{enumitem}
\usepackage[misc]{ifsym}
\usepackage[foot]{amsaddr}
\usepackage[hidelinks]{hyperref}
\usepackage{csquotes}
\usepackage[notref,notcite]{showkeys}
\usepackage[style=authoryear,ibidtracker=false,uniquename=false,giveninits=true,terseinits=true,maxbibnames=5,backend=biber]{biblatex}
\renewbibmacro{in:}{}
\addbibresource{rnni_convexity.bib}

@@

Since the problem of computing distances in the most common tree spaces as $\nni$ \autocite{Dasgupta2000-xa}, rooted $\spr$ \autocite{Bordewich2005-nx}, or $\tbr$ \autocite{Allen2001-ky} is known to be $\np$-hard, it is desirable to find a tree space that is less complex.
So far it is not known whether computing distances in $\rnni$ is $\np$-hard or not.
The natural approach to solve this question is to compare $\rnni$ and $\nni$ graph to see whether the proof of $\np$-hardness of $\nni$ applies to $\rnni$ as well.
We will explain why this is not the case by considering an example of shortest paths in $\nni$ that builds the key of the proof of its $\np$-hardness.
After explaining why this example cannot be transferred to $\rnni$ we will introduce the so-called Split Theorem (Conjecture~\ref{conjecture:split_theorem}), which has been conjectured in \autocite{Gavryushkin2018-ol}.
However, we present a counterexample disproving the Split Theorem in its original version and claiming the Cluster Theorem (Conjecture~\ref{conjecture:cluster_theorem}) as alternative version.

The proof of $\np$-completeness of $\nni$ in \autocite{Dasgupta2000-xa} is based on the fact that Theorem~\ref{thm:split_nni} holds for $\nni$.
The aim of this paper was to gain better understanding of the $\rnni$ space of ranked phylogenetic trees.
Before investigating geometrical properties of $\rnni$ we introduced some algorithms in Section~\ref{section:algorithms}).
One of the most important algorithms we introduced here is $\findpath$, which can be used for approximating $\rnni$ distances.
We could in particular show that this algorithm computes exact distances for all pairs of trees on up to six taxa.
This algorithm might be a useful tool for future investigations of the $\rnni$ space.

We used the algorithms introduced in Section~\ref{section:algorithms} to prove some geometric properties of the $\rnni$ graph such as diameter and radius.
Moreover, we discovered that the set of caterpillar trees is convex in $\rnni$ (Theorem~\ref{thm:caterpillar_convex}).
With this result and the algorithm $\csort$ for computing shortest caterpillar paths (Section~\ref{section:alg_csort}) we can compute $\rnni$ distances between caterpillar trees in polynomial time.

This already suggests that there is a difference in complexity of computing distances in $\rnni$ and $\nni$, since in $\nni$ there is no subspace known for which computing distances only takes polynomial time.
In Section~\ref{section:cluster_theorem} we supported the claim that computing $\rnni$ distances is not as hard as computing $\nni$ distances further by discussing the proof of $\np$-completeness of computing distances in $\nni$ and showing that this proof does not work for $\rnni$.
Specifically, we conjectured the Cluster Theorem (Conjetcure~\ref{conjecture:cluster_theorem}), which does not hold in $\nni$.
However, we strongly suspect that it holds in $\rnni$ and could prove that it does hold for trees on up to seven taxa \autocite{Collienne2019}.

\subsection{Open Problems}

In Section~\ref{section:cluster_theorem} we claimed that the Cluster Theorem holds for $\rnni$.
We could show that the previous version of it, the Split Theorem as it was conjectured in \autocite{Gavryushkin2018-ol}, is not true.
If the Cluster Theorem actually is true still remains an open question.
Proving it will be a big step towards finding the complexity of computing $\rnni$ distances.

This question of whether computing $\rnni$ distances is $\np$-hard is another problem for future research.
Though we did not aim to answer this question in this paper, we made some progress towards solving it by establishing some geometric properties of the $\rnni$ graph.
Comparing these results with what is known for the plain $\nni$ graph makes us assume that the complexity of computing shortest path in $\rnni$ might differ from $\nni$.
We also suspect that the fact that computing distances between caterpillar trees can be done in polynomial time might be the key for further investigation of the complexity of computing $\rnni$ distances.

\printbibliography

\end{document}

>>>>

\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage[UKenglish]{babel}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amssymb}
\usepackage{float}
\usepackage{graphicx}
\usepackage{algorithm}
\usepackage[noend]{algorithmic}
\usepackage{todonotes}
\usepackage{enumitem}
\usepackage[misc]{ifsym}
\usepackage[foot]{amsaddr}
\usepackage[hidelinks]{hyperref}
\usepackage{csquotes}
%\usepackage[notref,notcite]{showkeys}
\usepackage[style=authoryear,ibidtracker=false,uniquename=false,giveninits=true,terseinits=true,maxbibnames=5,backend=biber]{biblatex}
\renewbibmacro{in:}{}
\addbibresource{rnni_convexity.bib}

@@

The problem of computing distances in all common phylogenetic graphs, including $\nni$ \autocite{Dasgupta2000-xa}, $\spr$ \autocite{Bordewich2005-nx}, and $\tbr$ \autocite{Allen2001-ky}, is $\np$-hard.
Although $\np$-hardness does not necessarily imply practical impossibility, this so far has been the case for these phylogenetic algorithms with extremely few major advances \autocite{Whidden2010-bw}.
$\nni$ is especially known for its algorithmic hardness \autocite{Whidden2018-fw}.

Our results suggest a surprising advance in this field of $\np$-problems -- adding ranking information to trees simplifies some of the algorithms and makes some unintuitive counterexamples impossible.
We hence continued the investigation of the $\rnni$ graph on ranked phylogenetic trees in this paper.
We have taken the route of comparing the classic $\nni$ with the $\rnni$ graph, so have mostly been concentrating on questions that are settled in $\nni$.
Specifically, we have considered a number of geometric properties, including the diameter, radius, convexity of caterpillar trees, and the Split Theorem.
We have also justified that the technique of \textcite{Dasgupta2000-xa} to prove that distances in $\nni$ are $\np$-hard to compute cannot be directly applied in $\rnni$.
This led us to the Cluster Theorem, a statement that does not hold in $\nni$.
All our algorithms and methods developed in this paper witness in favour of the Cluster Theorem being true in $\rnni$.
For example, we have checked the statement computationally for up to seven taxa \autocite{Collienne2019}.
Although the Split Theorem is true in $\spr$, it is still $\np$-hard to compute distances in that graph.
Hence Cluster Theorem alone cannot be used as an argument in favour of the distance problem having an efficient solution in $\rnni$.
But the greedy nature of all our algorithms for computing shortest paths in $\rnni$ developed in this paper does provide such an argument.

The algorithms we developed and implemented in this paper have served as a main ingredient for our study of the geometry of $\rnni$ space.
These algorithms are of interest on their own as well.
For example, our $\findpath$ algorithm gives a good approximation of the $\rnni$ distance, performing exactly on small trees.
We are not aware of the minimal number of taxa where this algorithm fails to return the correct distance.
The $\mdtree$ algorithm produces trees as far away from a given tree as possible, a task of importance in simulation and model comparison studies.
The $\csort$ algorithm efficiently computes the $\rnni$ distance and shortest paths between caterpillar trees exactly.
To the best of our knowledge no such algorithm exists for $\nni$.
This implies that the algorithmic complexity of comping distances in $\nni$ is higher than in $\rnni$.

The question of whether computing $\rnni$ distances is $\np$-hard is still an open problem.

\section*{Acknowledgements}
We thank Charles Semple for useful discussions about the Cluster Theorem.

We acknowledge support from the Royal Society of New Zealand through the Rutherford Discovery Fellowship awarded to AG.
This work was partially supported by MBIE through the Endeavour Smart Ideas and Data Science Programmes grants.

Part of this work was done while LC and AG were visiting Max Plank Institute for Evolutionary Biology, we appreciate their support.

\printbibliography
\end{document}
